Gaia's constraints on the extremely massive ($\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-12 VFTS682_APJL.tex
new fig and angle data: TO BE CHECKED
text update: rewrote sec. 3, description of fig. 2 still drafty. Need the figure to improve, and I'm not convinced about the direction we are trying for it with Eva

\author{M.~Renzo$^{*,1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{4}$, 
  J.~Bestenlehner$^{5}$, \SdM{C.~J.~Evans$^{X}$},
  V.~H\'enault-Brunet$^{6}$,  S.~Justham$^{1,7,8}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{9}$,  \SdM{F. Najarro$^{X}$}, H.~Sana$^{10}$, F.~R.~Schneider$^{11}$, J.~S.~Vink$^{12}$}
\affil{ $^{*}$ Corresponding author.  A list of the affiliations can be found at the end of this paper.}

 \thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\begin{abstract}

@@

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      &\multirow{2}{*}{surface helium abundance} &  & 0.45 & (2)\\
      & &  & 0.49 & (3)\\
      \hline

@@

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\
      \hline
      \multicolumn{5}{l}{\emph{Coordinates}} \\
      \hline
         &$\mathrm{RA}_\mathrm{VFTS682}$&[degrees] & 84.7314 %$ \pm 0.036\,\mathrm{mas} $ 
                     & (1) \\        
               &$\mathrm{DEC}_\mathrm{VFTS682}$&[degrees] & -69.0741%$ \pm 0.048\,\mathrm{mas} $
                     & (1)  \\

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & 84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & 0.0547                      
        &  \SdM{your value? }
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] &0.0268 
                     &  \SdM{your value ? } \\

       \hline
           \multicolumn{5}{l}{\emph{Gaia's proper motion for VFTS 682
      and average of the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$ &[$\masyr$] & $0.026$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (2) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (2)\\
\hline

@@

      \multicolumn{5}{l}{\emph{Gaia's relative proper motion for VFTS
      682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,4) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,4) \\
      &$\delta\mu_\mathrm{}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,4) \\
                 &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,4)\\  
                     &$\theta$  &[degrees] &  $14.2_{-43}^{+44}$  & (1,4)\\  
                         &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,4)\\  
 %

 \hline     
          \multicolumn{5}{l}{\emph{HST's relative proper motion for VFTS 682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (3) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (3) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (3) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (3)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,4)\\  
                          &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,4)\\  
 %

@@

    \end{tabular}
    \tablecomments
    { Proper motions are based on Gaia data unless indicated otherwise in subscript.    The conversion factor assuming a distance of 50\,kpc is such that $1\masyr$ corresponds to $237\kms$.  The correlation coefficient is indicated as $\rho$. The angle $\theta$ is defined such that $\theta=0$ for motions pointing away radially from R136.
    (1) \cite{brown:18}, 
    (2) \cite{lennon:18}, 
    (3) \cite{platais:18} and
    (4) {\color{blue}This work}.
    %The error on the RA and DEC positions, are of order $\sim$$10^{-2}\masyr$ in \emph{Gaia} DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{Left panel: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. Right panel:  distribution of
    angles between the relative proper motion direction and the radial
    direction to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.}
  \label{fig:mu_dist}
\end{figure*}

@@

%\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e. groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.026, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

\SdM{Since the errors for HST and Gaia are so similar and yield to different directions, I think we should probably restructure this paragraph.  I liked how you break down the story in two separate questions, but given Fig 1 I think it may not work well.  The first question should have adressed it in context of the motions of the other stars. And it has a very typical motion. I think that instead, you should simply describe Fig 1 step by step.  I think that will be stronger.  }

\subsection{Is it a runaway star?}
\label{sec:runaway}
We first address the question whether VFTS682 is a typical star
from the kinematic point of view (which is what would be expected if it formed in relative isolation), or whether has a significant peculiar velocity compared to its surrounding population (which is what we would expect if the star has a runaway origin).

After subtracting the mean proper motion of the field, we obtain the relative two-dimensional relative proper motion. We obtain  $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$   and $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$.

Both indicate a non-zero peculiar 2D motion that would make it a slow runaway star, but in both cases the meaurements are also consistent with zero within 2 sigma.

Figure~\ref{fig:mu_dist} shows the distribution of relative proper motions for other VFTS O, B and WR stars subselecting either those with relative proper motion errors smaller than $0.1\masyr$ (yielding 317 stars) or a more restrictive cut including only those with proper motion errors smaller than $0.05\masyr$ (yielding 36) stars.   We see that the derived Gaia proper motion is very typical for stars in the field.

\SdM{XXX We need to rethink }

@@

.

@@

\subsection{Does it come from the R136 cluster?}
\label{sec:r136_origin}

@@

Figure~\ref{fig:mu_dist} shows the distribution in proper motion relative to R136
of OB-type and Wolf-Rayet stars included both in the VFTS survey and
\emph{Gaia DR2} with reliable astrometric solutions. The bottom panel
shows the distribution in angles $\theta$ between the relative proper motion
direction and the direction from the core of R136 to the star. We
emphasize that our subset of stars with reliable proper motion
measurements is biased towards the fast moving objects, which results in a
distribution of angles mildly peaked at small angles.

VFTS682 is not an outlier in relative proper motion%  compared
% to the two O-type runaways (re-)identified by \cite{lennon:18} (see
% also below)
. However, the star was expected to be a ``slow runaway'' by \citet{bestenlehner:11} in
the dynamical ejection scenario. The green shade in
\Figref{fig:mu_dist} shows the range of relative proper motions
required to reach the present day location within
the uncertainties on the apparent age of the star, assuming it was
ejected from R136 very early in its life.

Because of the large
error bars, the angle between the relative proper motion is not very
constraining, but it is suggestive that the best value is close to zero.
Therefore, the relative proper motion from \emph{Gaia} DR2 are consistent with the hypothesis
of dynamical ejection.

The red arrow in \Figref{fig:main} shows the direction of relative proper motion of
VFTS682 from \emph{Gaia} DR2, and the
yellow arrows illustrate the uncertainty. % show the possible
% range of directions within the uncertainties in the measured relative proper
% motion.
These arrows cross at the present-day location of VFTS682 and
are prolonged in the direction opposite to the motion to illustrate
the possible range of origins. % The yellow arrow shows the direction of
% the relative proper motion from HST (see \Eqref{eq:pm_around_HST}),
% but we omit to show the possible range of directions from HST data, since it is
% insufficiently constrained to be useful.

@@

The kinematic age of this star, assuming it originates from R136, is

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu^\mathrm{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age
of $1.0\pm 0.2$\,Myr \citep{schneider:18}. % , which
% corroborates the idea that the star is the result of a dynamical
% ejection very soon in the cluster evolution.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional speed of
$44\pm21\,\kms$. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

@@

The variability of VFTS682, reminiscent of LBV stars, suggests
that VFTS682 (and therefore its analogs in the core of R136) might
experience enhanced mass loss episodes in LBV eruptions. \citet{smith:15} made the highly
debated\footnote{See, e.g., \cite{humphreys:16, davidson:16, smith:16}.}
claim that LBV stars are typically isolated form O-type stars. The fact that VFTS682 is a dynamically
ejected runaway which might evolve into an LBV star suggests that
N-body interactions also play a role in explaining the apparent
isolation of at least some LBV stars.

\citet{lennon:18} carried out a study similar to ours on the fast
moving O-type stars
in the region, and found two massive runaway stars
($\sim$$90\,M_\odot$) in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS 72) is less clear given its direction
of motion. The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 (see \Secref{sec:r136_origin}) is smaller than the
corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic
age of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

>>>>

\author{M.~Renzo$^{*,1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{4}$, 
  J.~Bestenlehner$^{5}$, C.~J.~Evans$^{6}$,
  V.~H\'enault-Brunet$^{7}$,  S.~Justham$^{1,8,9}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{10}$, F. Najarro$^{11}$, H.~Sana$^{12}$, F.~R.~Schneider$^{13}$, J.~S.~Vink$^{14}$}
\affil{ $^{*}$ Corresponding author, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}.  A list of the affiliations can be found at the end of this paper.}

\begin{abstract}

@@

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      &\multirow{2}{*}{surface helium mass fraction} &  & 0.45 & (2)\\
      & &  & 0.49 & (3)\\
      \hline

@@

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\
      \hline
      \multicolumn{5}{l}{\emph{Coordinates}} \\
      \hline
         &$\mathrm{RA}_\mathrm{VFTS682}$&[degrees] & \phantom{-}84.7314 %$ \pm 0.036\,\mathrm{mas} $ 
                     & (1) \\        
               &$\mathrm{DEC}_\mathrm{VFTS682}$&[degrees] & -69.0741%$ \pm 0.048\,\mathrm{mas} $
                     & (1)  \\

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & \phantom{-}84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & \phantom{-0}0.0547                      
        &  (2, 5)
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] & \phantom{-0}0.0268 
                     &  (2, 5) \\

       \hline
           \multicolumn{5}{l}{\emph{Gaia proper motion for VFTS682
      and average of the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$ &[$\masyr$] & $0.026$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (3) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (3)\\
\hline

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta$  &[degrees] &  $14.2_{-43}^{+44}$  & (1,5)\\  
      &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                          &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

@@

    \end{tabular}
    \tablecomments
    { Proper motions are based on Gaia data unless indicated otherwise in subscript.    The conversion factor assuming a distance of 50\,kpc is such that $1\masyr$ corresponds to $237\kms$.  The correlation coefficient is indicated as $\rho$. The angle $\theta$ is defined such that $\theta=0$ for motions pointing away radially from R136.
      (1) \cite{brown:18},
      (2)~\cite{bestenlehner:11},
    (3) \cite{lennon:18}, 
    (4) \cite{platais:18} and
    (5) {\color{blue}this study}.
    %The error on the RA and DEC positions, are of order $\sim$$10^{-2}\masyr$ in \emph{Gaia} DR2. %REPETITION: %  We adopt the peculiar
      % radial velocity calculated using the  NV $\lambda4944$ line, which is
      % less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy}}
  \label{fig:dist}
\end{figure*}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e. groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.026, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light shaded
blue). We highlight VFTS682 together with the two O-type
runaways \mbox{(re-)identified} by \cite{lennon:18} (VFTS16 and VFTS72,
respectively). We emphasize that selecting stars with small error bars
implicitly biases the sample toward fast moving objects.

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star}).

The right panel shows the distributions in flight direction for these
stars: $\theta$ is the angle between the radial direction from the
center of R136 to each star and its proper motion relative to
R136 (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. \todo{fix colors} The dashed circle shows the location of the cluster, and the red
(cyan) arrow shows the relative proper motion from \emph{Gaia} DR2
(HST). The lighter arrows show the corresponding opening angle within
the large uncertainties. The length of each arrow is proportional to
the speed of the star, and dashed prolungations show the possible
region of origin assuming an age of $1.0\pm0.2$\,Myr
\citep[][]{schneider:18}. This figure illustrates that R136 is clearly
the most likely origin of the star, although the large error bars
prevent a robust identification.

Assuming VFTS682 originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

Therefore, although the motion of VFTS682 is not an outlier compared
to other massive stars, it is possible to tentatively claim that it is
a slow runaway with velocity compatible with dynamical ejection from
R136 early in the evolution of both the star and the cluster.

% \subsection{Is it a runaway star?}
% \label{sec:runaway}
% We first address the question whether VFTS682 is a typical star
% from the kinematic point of view (which is what would be expected if it formed in relative isolation), or whether has a significant peculiar velocity compared to its surrounding population (which is what we would expect if the star has a runaway origin).

% After subtracting the mean proper motion of the field, we obtain the relative two-dimensional relative proper motion. We obtain  $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$   and $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$.

% Both indicate a non-zero peculiar 2D motion that would make it a slow runaway star, but in both cases the meaurements are also consistent with zero within 2 sigma.

% Figure~\ref{fig:mu_dist} shows the distribution of relative proper motions for other VFTS O, B and WR stars subselecting either those with relative proper motion errors smaller than $0.1\masyr$ (yielding 317 stars) or a more restrictive cut including only those with proper motion errors smaller than $0.05\masyr$ (yielding 36) stars.   We see that the derived Gaia proper motion is very typical for stars in the field.

% \SdM{XXX We need to rethink }

@@

@@

% \subsection{Does it come from the R136 cluster?}
% \label{sec:r136_origin}

@@

% Figure~\ref{fig:mu_dist} shows the distribution in proper motion relative to R136
% of OB-type and Wolf-Rayet stars included both in the VFTS survey and
% \emph{Gaia DR2} with reliable astrometric solutions. The bottom panel
% shows the distribution in angles $\theta$ between the relative proper motion
% direction and the direction from the core of R136 to the star. We
% emphasize that our subset of stars with reliable proper motion
% measurements is biased towards the fast moving objects, which results in a
% distribution of angles mildly peaked at small angles.

% VFTS682 is not an outlier in relative proper motion%  compared
% % to the two O-type runaways (re-)identified by \cite{lennon:18} (see
% % also below)
% . However, the star was expected to be a ``slow runaway'' by \citet{bestenlehner:11} in
% the dynamical ejection scenario. The green shade in
% \Figref{fig:mu_dist} shows the range of relative proper motions
% required to reach the present day location within
% the uncertainties on the apparent age of the star, assuming it was
% ejected from R136 very early in its life.

% Because of the large
% error bars, the angle between the relative proper motion is not very
% constraining, but it is suggestive that the best value is close to zero.
% Therefore, the relative proper motion from \emph{Gaia} DR2 are consistent with the hypothesis
% of dynamical ejection.

% \todo{discuss and find a place for Fig.2}
% The red arrow in \Figref{fig:main} shows the direction of relative proper motion of
% VFTS682 from \emph{Gaia} DR2, and the
% yellow arrows illustrate the uncertainty. % show the possible
% % range of directions within the uncertainties in the measured relative proper
% % motion.
% These arrows cross at the present-day location of VFTS682 and
% are prolonged in the direction opposite to the motion to illustrate
% the possible range of origins.
% The yellow arrow shows the direction of
% the relative proper motion from HST (see \Eqref{eq:pm_around_HST}),
% but we omit to show the possible range of directions from HST data, since it is
% insufficiently constrained to be useful.

@@

 % , which
% corroborates the idea that the star is the result of a dynamical
% ejection very soon in the cluster evolution.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar three-dimensional speed of
$44\pm21\,\kms$ \todo{make 2d? use LOS velocity from Marimar?}. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

@@

The variability of VFTS682, reminiscent of LBV stars, suggests
that VFTS682 (and therefore its analogs in the core of R136) might
experience enhanced mass loss episodes in LBV eruptions. \citet{smith:15} made the highly
debated\footnote{See, e.g., \cite{humphreys:16, davidson:16, smith:16}.}
claim that LBV stars are typically isolated from O-type stars. The fact that VFTS682 is a dynamically
ejected runaway which might evolve into an LBV star suggests that
N-body interactions also play a role in explaining the apparent
isolation of at least some LBV stars.

\citet{lennon:18} carried out a study similar to ours on the fast
moving O-type stars
in the region, and found two very massive runaway stars in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS 72) is less clear given its direction
of motion (cf.~\Figref{fig:dist}). The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 is smaller than the
corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic
age of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.
